The Concurrent Calculi Formalisation Benchmark

### 2024-02-19 short-paper/short-paper.tex
linearity section revisited

Our benchmark problems (detailed in \cref{app:challenges}) consider
\emph{in isolation} three key aspects that frequently pose challenges
when mechanising concurrency theory: \emph{linearity}, \emph{scope
  extrusion}, and \emph{coinductive reasoning}.  Mechanisations must
often address several of these aspects at the same time; however, we
see the combination of techniques as a next
step%beyond the scope of this challenge
, as discussed further in \Cref{sec:going-beyond}.  Our benchmark
challenges are based on process calculi, as these provide a simple but
realistic showcase of common issues encountered during mechanisation
efforts.
% While the three highlighted aspects are not the sole sources of difficulties, they emerge in most mechanisations of concurrency theory.

We have begun collecting solutions to our challenges on our website:
%
\begin{center}
  \url{https://concurrentbenchmark.github.io/}
\end{center}
%
In the longer term, we expect to use the website for promoting best
practices, tutorials and guidelines derived from solutions to our
challenges.  We encourage anyone interested to try the challenges
using their favourite tools and techniques, and to send us their
solutions.

\section{The Benchmark and its Design}\label{sec:design-discussion}
\alb{Next paragraph needs rewriting}
In this section, we outline the factors considered in designing the benchmark challenges.
We begin with some general remarks, then describe the individual design considerations for each challenge problem, and the evaluation criteria for solutions.

Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
% \alb{Should we stress we are interested in the \textbf{meta-theory} of process calculi, not}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}
%
Note that the challenges revolve around the mechanization of the
\emph{meta-theory} of process calculi, rather than interpreters or
type checkers for the latter.

To provide a framework in which to answer these questions, our benchmark is designed to satisfy three main goals:
\begin{enumerate}[label=\textbf{(G\arabic*)},leftmargin=10mm]
\item\label{item:goal-comperison-accessibility} To enable the comparison of
  proof mechanisation approaches, making the challenges accessible to
  mechanisation experts who may be unfamiliar with concurrency theory;

@@

\todo{Refer to the future benchmark section about other things one could do, but they still all need the basics}
Still, these (and other) applications need the basic
techniques that our challenge problems emphasise:
Aligned with design goal~\ref{item:goal-reusability}, our problems are rooted in the fundamental meta-theory of process calculi, without necessitating the creation of novel theory. These problems centre around well-established theorems and results, showcasing interesting proof techniques that can be leveraged in future endeavours.
To achieve design goal \ref{item:goal-comperison-accessibility}, we have
formulated our challenge problems to explore the three aspects
(linearity, scope extrusion, coinduction) independently, so that they
may be solved individually and in any order; each problem is
reasonably small and easily understandable with basic knowledge of
textbook concurrency theory, process calculi, and type theory.  The
process calculus used in each challenge focuses on the features that
we want to emphasise, and omits all constructs (such as choices) that
would complicate the mechanisation without bringing additional
tangible benefits and insights.  The minimality and uniformity of the
calculi also allows us to target design goal
\ref{item:goal-tutorials}.

\subsubsection{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.

\cite{Brady2017} notes that using algebraic effects is unwieldy when there are many communication channels and suggests an experimental approach using dependent uniqueness types.
\cite{Zalakain2019} notes that systems lacking linear types must simulate linearity, and that using a linearity predicate complicates congruences.
They had to limit themselves to an asymmetric structural precongruence to be able to prove subject reduction.

\alb{We need more details about what the challenge is}
For our challenge on linear reasoning we have chosen a type safety theorem
for a system using session types, combining as usual type preservation with absence of errors.
This allows us to introduce linear reasoning with few definitions.
With our objective focused on linearity, we forego channel delegation
and opt for a reduction approach rather than employing a transition
system semantics.  Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names, highlighting
their duality within the type system.  We have chosen a simple notion
of well-formedness, and do not  consider
deadlocks. While more sophisticated notions of well-formedness are
interesting, the proofs of these properties would distract from the
core linear aspects of the challenge.

While a linear context is perhaps best seen as a multi-set, most proof assistants have better support for list processing.
The latter representation is intuitive, but may require establishing a large number of technical lemmata that are orthogonal to the mathematics of the problem under study, say proving type preservation for session types.
In this space, several designs are possible: one can label occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}) or \alb{something --- ask Francisco} as in \cite{Castro2020}.
Some authors instead impose a multiset structure over lists (\eg~\cite{ChaudhuriLR19,Danielsson12}); still, the effort required to develop the infrastructure is significant.

@@

\alb{This needs some work: we do not describe the challenge at all and the citations are not the one I had in mind}
\subsubsection{Scope extrusion.}
The encoding of binding constructs has been extensively studied
w.r.t.\ the $\lambda$-calculi, but process calculi present additional
challenges.  $\pi$-calculi typically include several different binding
constructs: inputs bind a received name or value, recursive processes
bind recursion variables, and restrictions bind names.  The first two
act similarly to the binders known from $\lambda$-calculi, but
restrictions may be more challenging due to scope extrusion.

>>>>

Our benchmark problems (detailed in \cref{app:challenges} and on the website) consider
\emph{in isolation} three key aspects that frequently pose challenges
when mechanising concurrency theory: \emph{linearity}, \emph{scope
  extrusion}, and \emph{coinductive reasoning}.  Mechanisations must
often address several of these aspects at the same time; however, we
see the combination of techniques as a next
step%beyond the scope of this challenge
, as discussed further in \cref{sec:going-beyond}.  Our benchmark
challenges are based on process calculi, as these provide a simple but
realistic showcase of common issues encountered during mechanisation
efforts.
% While the three highlighted aspects are not the sole sources of difficulties, they emerge in most mechanisations of concurrency theory.

We have begun collecting solutions to our challenges on our website:
%
\begin{center}
  \url{https://concurrentbenchmark.github.io/}
\end{center}
%
In the future, we expect to use the website to promote best practices and tutorials derived from solutions to our challenges.
We encourage everyone to try the challenges using their favourite techniques, and to send us their solutions.

\section{The Benchmark and its design}\label{sec:design-discussion}
\alb{Next paragraph needs rewriting}
In this section, we outline the factors considered in designing the benchmark challenges.
We begin with some general remarks, then describe the individual design considerations for each challenge problem, and the evaluation criteria for solutions.

Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
% \alb{Should we stress we are interested in the \textbf{meta-theory} of process calculi, not}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in tools for mechanising the meta-theory of process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting new mechanisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}
%
Note that the challenges revolve around the mechanization of the
\emph{meta-theory} of process calculi, rather than interpreters or
type checkers for the latter.

To provide a framework in which to answer these questions, our benchmark is designed to satisfy three main design goals:
\begin{enumerate}[label=\textbf{(G\arabic*)},leftmargin=10mm]
\item\label{item:goal-comperison-accessibility} To enable the comparison of
  proof mechanisation approaches by making the challenges accessible to
  mechanisation experts who may be unfamiliar with concurrency theory;

@@

To achieve design goal \ref{item:goal-comperison-accessibility}, we have
formulated our challenge problems to explore the three aspects
(linearity, scope extrusion, coinduction) independently, so that they
may be solved individually and in any order; each problem is
reasonably small and easily understandable with basic knowledge of
textbook concurrency theory, process calculi, and type theory.  The
process calculus used in each challenge focuses on the features that
we want to emphasise, and omits all constructs (such as choices) that
would complicate the mechanisation without bringing additional
tangible insights.  The minimality and uniformity of the
calculi also allows us to target design goal
\ref{item:goal-tutorials}.
Aligned with design goal~\ref{item:goal-reusability}, our problems are rooted in the fundamental meta-theory of process calculi, without necessitating the creation of novel theory.
Our challenges centre around well-established theorems and results, showcasing interesting proof techniques that can be leveraged in many applications (as we will further discuss in \cref{sec:going-beyond}).

\subsubsection{Linearity.}

% \cite{Brady2017} notes that using algebraic effects is unwieldy when there are many communication channels and suggests an experimental approach using dependent uniqueness types.
% \cite{Zalakain2019} notes that systems lacking linear types must simulate linearity, and that using a linearity predicate complicates congruences.
% They had to limit themselves to an asymmetric structural precongruence to be able to prove subject reduction.

For our challenge on linear reasoning we have chosen a type safety
theorem for a system using session types, combining as usual subject reduction with absence of errors.  This allows us to introduce
linear reasoning with few definitions.  With this focus, we forego
channel delegation: input and output receives/sends only values, which
do not include channels, hence the topology of the communication
network described by a process cannot change.  We do not allow
recursion or replication in the syntax, hence no infinite behaviours
can be expressed. In particular, we only model linear (as opposed to
shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names, highlighting their
duality within the type system. The latter enforces
 two invariants, namely no endpoint is used simultaneously by parallel processes and
 the two endpoints of the same session have dual types.
The operational semantics is formulated in term of reductions rather
than via a transition system, allowing implementors to explore
alternative ways to encode structural congruence\footnote{\alb{Should we mention this in challenge 2?}}.

The statement of subject reduction is the usual one, encompassing type
preservation for structural congruence. W.r.t.\ type safety, we have
chosen a simple notion of what an error is: well-typed processes are a
well-formed in the sense that they do not use endpoints in a non-dual
way.  While more sophisticated notions of well-formedness that
prevents deadlocks are interesting, the proofs of these properties
would distract from the core linear aspects of the challenge.

In mechanised meta-theory, addressing linearity means choosing an
appropriate representation of the linear context.  The crux of the
matter is \emph{context splitting}, as showcased by the typing rule
for parallel composition.  While a linear context is perhaps best seen
as a multi-set, most proof assistants have better support for list
processing.  The latter representation is intuitive, but may require
establishing a large number of technical lemmata that are orthogonal
to the mathematics of the problem under study, in our case proving
type safety for session types.  In this space, several designs are
possible: one can label occurrences of resources to constrain their
usage (\eg~\cite{CicconeP20}); some authors instead impose a multiset
structure over lists
(\eg~\cite{Danielsson12,ChaudhuriLR19}). Alternatively, contexts can
be implemented as finite maps (\cite{Castro2020}), whose operations
are sensitive to a linear discipline. In all these cases, the effort
required to develop the infrastructure is significant.

@@

\subsubsection{Scope extrusion.}
\alb{This needs some work: we do not describe the challenge at all and the citations are not the one I had in mind}

The encoding of binding constructs has been extensively studied
w.r.t.\ the $\lambda$-calculi, but process calculi present additional
challenges.  $\pi$-calculi typically include several different binding
constructs: inputs bind a received name or value, recursive processes
bind recursion variables, and restrictions bind names.  The first two
act similarly to the binders known from $\lambda$-calculi, but
restrictions may be more challenging due to scope extrusion.
